
# Regression

```{r, include=FALSE}
library(tidyverse)
library(kableExtra)
knitr::opts_chunk$set(echo = TRUE, fig.width=3, fig.height=2, fig.align = "center",
                      message=FALSE, warning=FALSE)
doc_theme <- ggplot2::theme_bw()
```

## Slope and Intercept

Take a look at the scatterplot below. It shows several potential lines to describe the relationship between x and y. Indeed, each of these lines seems reasonable. We can imagine many more reasonable lines.

```{r echo=FALSE, fig.height=2, fig.width=3, message=FALSE, warning=FALSE}
set.seed(34984)
df <- tibble(x = c(rnorm(25), 5), 
             y = x + c(rnorm(25), -5))

ggplot(df, aes(x = x, y = y)) + 
  geom_point() + 
  geom_smooth(se = FALSE, method = "lm", color = "grey50", size = 0.5) + 
  geom_smooth(se = FALSE, method = MASS::rlm, color = "grey50", size = 0.5) + 
  geom_smooth(se = FALSE, method = MASS::rlm, method.args = list(psi = MASS::psi.hampel), color = "grey50", size = 0.5) + 
  geom_smooth(se = FALSE, method = MASS::rlm, method.args = list(psi = MASS::psi.bisquare), color = "grey50", size = 0.5) + 
  geom_smooth(se = FALSE, method = robustbase::lmrob, method.args = list(), color = "grey50", size = 0.5) + 
  geom_smooth(se = FALSE, method = robustbase::lmrob, method.args = list(method = "D"), color = "grey50", size = 0.5) + 
  theme_minimal()
```

But what is the **best** line? 

Least squares offers *one* principle way to choose the best line.

The least squares principles says to choose the line that minimizes the RMS of of the residuals.

### Multiple Regression

